Data analysis to prioritise new cycleways in response to COVID-19

### 2020-06-03 report.Rmd
Update report

---
title: "Methods to prioritise pop-up active transport infrastructure"
# : overview and findings from 10 UK cities"
author: "Robin Lovelace, Malcolm Morgan, Joey Talbot, Martin-Lucas-Smith"
# output: bookdown::github_document2
# output: bookdown::html_document2
# output:
#   bookdown::pdf_document2:
#     keep_tex: true
# bibliography: references.bib
bibliography: ref.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE, cache = FALSE, out.width = "100%")
```

```{r, echo=FALSE, eval=FALSE}
citr::tidy_bib_file("article.Rmd", "~/uaf/allrefs.bib", file = "ref.bib")
wordcountaddin::word_count("article.Rmd")
knitr::purl("article.Rmd")
file.copy("article.R", "code/article-methods-1.R")
pagedown::chrome_print("article.html")
browseURL("article.pdf")
zf = list.files(pattern = "tex|png", full.names = TRUE)
zfl = list.files("article_files/figure-latex/", full.names = TRUE)
fz = c(zf, zfl)
msgf = paste(fz, collapse = " ")
msgf = gsub(pattern = "\\./", replacement = "", msgf)
msg = paste("tar -cvf article.tar ", msgf)
system(msg)
```

# Abstract

<!-- This paper reports on methods developed to support the identification of pop-up active transport infrastructure. -->
In the context of reduced public transport capacity in the wake of the COVID-19 pandemic, governments are scrambling to enable walking and cycling.
A range of pop-up options exist.
The focus of this article is lane reallocation, which represents a 'quick win' for cities with roads that have a spare lane during reduced motor traffic conditions.
We found that the methods could condense the complexity of cities down to the most promising roads, which match intuition.
The evidence resulting from the methods, and future refinements, could support more evidence-based use of resources that have been made available to support implementation of pop-up schemes.

<!-- A range of pop-up options exists, including re-allocating road space, preventing through motor traffic and complete pedestrianisation of roads. -->
# RESEARCH QUESTIONS AND HYPOTHESIS

<!-- Currently the intro is 350 words, just about OK! -->

<!-- The COVID-19 pandemic has transformed all sectors of the economy, not least transport. -->
<!-- [Demand for long distance trips has plummetted](https://osf.io/preprints/socarxiv/v3g5d/) and [airline companies have gone bust](https://arxiv.org/abs/2004.08460) [@iacus_estimating_2020; @jittrapirom_exploratory_2020]. -->
Much attention has focused on the impacts of COVID-19 on long-distance travel patterns [e.g. @iacus_estimating_2020; @jittrapirom_exploratory_2020].
Yet short distance travel patterns have also changed, with a notable increase in active travel, particularly cycling, in some areas  [@harrabin_boom_2020].
The two main explanations for this are 1) the need increased need for exercise close to home during lockdowns for mental and physical health [@jimenez-pavon_physical_2020], and 2) a reduction in both public transport options and use [e.g. @tian_investigation_2020].
The second reason is particularly important given that many 'key workers' are low paid, with limited access to private automobiles.

<!-- due to cuts in services and fear of being infected while in enclosed spaces, meaning that walking and particularly cycling may be the only way that workers can reach key destinations such as hospitals. -->
<!-- From a physical activity perspective this change is welcome: obesity is a powerful predictor of all-cause mortality, including from COVID-19 [@docherty_features_2020]. -->
<!-- From a physical distancing perspective, increased levels of walking and cycling, -->
<!-- the shift creates pressure on governments to ensure sufficient 'space for social distancing', while enabling increased active mobility for health and travel to key workplaces. -->
<!-- particularly in densly populated urban areas where pavements and cycleways may be narrow,  -->

Local and national governments are working out how best to respond.
Many options are available to ensure that citizens can benefit from outdoor activity while minimising health risks, ranging from the hand sanitiser provision to the creation of extra active transport space [@freeman_covid19_2020].
Installation of 'pop-up' active transport infrastructure has been endorsed and implemented in many places [@laker_world_2020].
The Scottish government, for example, has provided £10 million "to keep key workers moving" by "reallocating road space to better enable this shift and make it safer for people who choose to walk, cycle or wheel for essential trips or for exercise" [@transportscotland_10_2020].
On 9^th^ May 2020, the UK government announced a £250 million package for pop-up active transport infrastructure [@reid_government_2020].
Significantly, alongside this funding comes updated [guidance](https://www.gov.uk/government/publications/reallocating-road-space-in-response-to-covid-19-statutory-guidance-for-local-authorities/traffic-management-act-2004-network-management-in-response-to-covid-19) on pop-up infrastructure and safety [@govuk_transport-secretary_2020].
Evidence is needed to ensure that such investment is spent effectively and where it is most needed.

Most pop-up active transport infrastructure can be classified into three broad categories:

1. 'filtered permeability', e.g. as [shown](https://twitter.com/CatrionaSwanson/status/1258322956595453952) in [@salfordcitycouncil_salford_2020]

2. banning cars and to pedestrianise streets, as in New York's 'Open Streets' scheme [@litman_pandemicresilient_2020]
3. the reallocation of one or more lanes on wide roads to create pop-up cycleways and pavements [@orsman_covid_2020].
<!-- interventions to prevent through traffic, with interventions as part of Salford's ['Liveable Streets' project](https://salfordliveablestreets.commonplace.is/) being a prominent example  -->
<!-- (see [here](https://twitter.com/CatrionaSwanson/status/1258322956595453952) for a photo illustrating this type of intervention) -->

The focus of this article is on the third category.
The research question is:

> How can automated data analysis and interactive visualisation methods help prioritise the reallocation of road space for pop-up active transport infrastructure?

Because of the recent, localised and often ad-hoc nature of pop-up infrastructure, it is difficult to make, let alone test, hypotheses related to the research question.
Our broad hypothesis is that digital tools based on open data, and crowdsourcing such as the interactive map used to support community-level responses to COVID-19 in Salford [@salfordcitycouncil_salford_2020], illustrated in Figure 1, can lead to more effective use of resources allocated to pop-up interventions.

```{r commonplace, fig.cap="Screenshot from the website salfordliveablestreets.commonplace.is to support local responses to the COVID-19 pandemic, including the prioritisation of pop-up active transport infrastructure."}
# knitr::include_graphics("https://user-images.githubusercontent.com/1825120/81451234-ed82d200-917b-11ea-977d-fff1665378c5.png")
# download.file("https://user-images.githubusercontent.com/1825120/81451234-ed82d200-917b-11ea-977d-fff1665378c5.png", "saferstreets.png")
knitr::include_graphics("saferstreets.png")
```

 <!-- With the rush to act, there is a great need for evidence of *where* new interventions should be prioritised. As with the medical science, research is needed now. Methods developed to identify locations of high walking and cycling potential can help ensure that the 'pop-up' infrastructure that goes in now is effective, safe, and placed where it is most needed. -->

# METHODS AND DATA

```{r load}
library(sf)
library(tidyverse)
library(tmap)
region_name = "West Yorkshire"
source("code/build.R")
if(!file.exists("rsf_leeds.Rds")) {
  u = "https://github.com/cyipt/popupCycleways/releases/download/0.1/rsf_leeds.Rds"
  download.file(u, "rsf_leeds.Rds")
}
r_original = readRDS("rsf_leeds.Rds")
if(!file.exists("rtid.csv")) {
  ur = "https://github.com/cyipt/popupCycleways/releases/download/0.1/rtid.csv"
  download.file(ur, "rtid.csv")
}
rtid = readr::read_csv("rtid.csv")
# uh = "http://media.nhschoices.nhs.uk/data/foi/Hospital.csv"
# download.file(uh, "uh.csv")
h = readr::read_delim("uh-edited.csv", delim = "|")
# nrow(h) # 1221 hospitals
h_clean = h %>% 
  mutate(Longitude = as.numeric(Longitude)) %>% 
  mutate(Latitude = as.numeric(Latitude)) %>% 
  filter(!is.na(Longitude)) %>% 
  filter(!is.na(Latitude))
hsf = h_clean %>% st_as_sf(coords = c("Longitude", "Latitude"), crs = 4326)
region = ukboundaries::leeds
tmap_mode("plot")
```

```{r parameters}
regexclude = "welling"
min_cycling_potential = 0
min_grouped_cycling_potential = 100
min_grouped_length = 500
city_centre_buffer_radius_small = 5000
city_centre_buffer_radius = 8000
city_centre_buffer_radius_large = 10000
key_destination_buffer_radius = 5000
city_centre = tmaptools::geocode_OSM("leeds", as.sf = TRUE)
city_centre_buffer_small = stplanr::geo_buffer(city_centre, dist = city_centre_buffer_radius_small)
city_centre_buffer = stplanr::geo_buffer(city_centre, dist = city_centre_buffer_radius)
city_centre_buffer_large = stplanr::geo_buffer(city_centre, dist = city_centre_buffer_radius_large)
h_leeds = hsf[city_centre_buffer, ]
h_leeds_buffer = stplanr::geo_buffer(h_leeds, dist = key_destination_buffer_radius)
city_key_buffer = st_union(city_centre_buffer, st_union(h_leeds_buffer))
```

```{r preprocess}
# remove motorways
r_original$highway_type = r_original$highway
r_original$highway_type = gsub(pattern = "_link", replacement = "", r_original$highway)
highway_table = table(r_original$highway)
highway_rare = highway_table[highway_table < nrow(r_original) / 100]
highway_remove = names(highway_rare)[!grepl(pattern = "motor|living|ped", x = names(highway_rare))]
r_cleaned = r_original %>% 
  filter(!grepl(pattern = "motorway", x = highway)) %>% 
  mutate(highway_type = case_when(
  highway_type %in% highway_remove ~ "other",
  grepl(pattern = "path|track|trunk", highway_type) ~ "other",
  grepl(pattern = "ped|liv", highway_type) ~ "pedestrian/living_street",
  TRUE ~ highway_type
  )) %>% 
  mutate(maxspeed = case_when(
    maxspeed <= 20 ~ "20 mph or less",
    maxspeed > 20 & maxspeed <= 30 ~ "30 mph",
    maxspeed > 30 ~ "40+ mph",
  )) %>% 
  mutate(cycling_potential = as.numeric(pctgov)) %>% 
  select(name, ref, highway, maxspeed, pctgov, width, Existing, length, highway_type, cycling_potential, rtid, idGlobal)
# table(r_cleaned$maxspeed)
rj = inner_join(r_cleaned, rtid) %>% 
  mutate(lanes_f = abs(lanespsvforward) + abs(lanesforward)) %>% 
  mutate(lanes_b = abs(lanespsvbackward) + abs(lanesbackward)) %>% 
  mutate(n_lanes_numeric = lanes_f + lanes_b) %>% 
  mutate(n_lanes = as.character(n_lanes_numeric))
rj$n_lanes_numeric[rj$n_lanes_numeric <= 0] = 1 
rj$n_lanes[rj$n_lanes == "0"] = "1"
rj$n_lanes[rj$n_lanes == "4"] = "4+"
rj$n_lanes[rj$n_lanes == "5"] = "4+"
rj$n_lanes[rj$n_lanes == "6"] = "4+"
cy = r_cleaned %>% filter(highway == "cycleway")
r_central = rj[city_key_buffer, ]
r_main = r_central %>% 
  filter(grepl(pattern = "cycleway|primary|second|tert", highway_type))
r_main_region = rj %>% 
  filter(grepl(pattern = "cycleway|primary|second|tert", highway_type))
```

Three key datasets were used for the project:

- Estimates of cycling potential to work at the street segment level from the UK Department for Transport funded Propensity to Cycle Tool (PCT) project [@goodman_scenarios_2019; @lovelace_propensity_2017]
- Data derived from OpenStreetMap, with several new variables added to support cycling infrastructure planning (see www.cyipt.bike for an overview)
- A list of hospital locations from the UK's National Health Service website [www.nhs.uk](https://www.nhs.uk/about-us/nhs-website-datasets/)
<!-- - Data on the location of road traffic casualties from -->

Datasets from the PCT and CyIPT project were merged, resulting in crucial variables summarised in Table 1.
A map showing the spatial distribution of hospitals in the case study city of Leeds is shown in Figure \@ref(fig:hospitals).

Table 1: Summary of the road segment dataset for Leeds

```{r t1, results='asis'}
# rj %>%
#   st_drop_geometry() %>%
#   # select(name, highway_type, maxspeed, cycling_potential, width) %>%
#   table1::table1(~ highway_type + cycling_potential + width + n_lanes | maxspeed, data = ., )
# ut = "https://user-images.githubusercontent.com/1825120/81476961-c32d2500-920c-11ea-8430-94c3afc2e27d.png"
# knitr::include_graphics(ut)
# system(paste("wget ", ut))
knitr::include_graphics("81476961-c32d2500-920c-11ea-8430-94c3afc2e27d.png")
```

```{r hospitals, fig.cap="Overview map of input data, showing the main highway types and location of hospitals in Leeds"}
m1 = r_main_region %>%
  mutate(`Highway type` = highway_type) %>% 
  mutate(`Cycling potential` = case_when(cycling_potential < 100 ~ 100, TRUE ~ cycling_potential)) %>% 
  tm_shape(bbox = st_bbox(city_centre_buffer)) +
  tm_lines(col = "Highway type", palette = c("green", "black", "blue", "grey"),
           lwd = "Cycling potential", scale = 5, lwd.legend = c(100, 200, 500, 1000),
           lwd.legend.labels = c("0 - 100", "100 - 200", "200 - 500", "500+")) +
  tm_shape(h_leeds) + tm_dots(size = 0.5, col = "ParentName", palette = "Dark2", title = "Hospital group") +
  tm_layout(legend.outside = TRUE)
# + tm_text("OrganisationName")
m1
```

<!-- Could say more about the case study city here if there is space -->

## Geographic subsetting

<!-- The region of analysis may seem like a basic consideration: most cities have well-defined administrative zones. -->
<!-- In Leeds and many other cities, it makes sense to focus on the region directly surrounding the city centre, in a kind of 'geographical triage' to omit from the analysis pop-up options in the outskirts and focus valuable attention on the routes that are most likely to serve the highest number of people. -->
We set a modifiable parameter `city_centre_buffer_radius` with an initial value of 8 km (5 miles) to geographically subset potential routes. 5 miles represents a distance that most people have the physical ability to cycle.
Figure \@ref(fig:gsub) shows the result of subsetting based on physical distance from the centre vs plotting all possible transport network segments within the city boundaries.
To ensure roads that could serve key destinations were included, the parameter `key_destination_buffer_radius` (initially set to 5 km) was used.
<!-- could say more... -->

```{r gsub, fig.cap="Illustration of geographic subsetting based on distance to a central point (Leeds city centre in this case) rather than based on location within somewhat arbitrarily shaped city boundaries. Radii of 5 km, 8 km and 10 km are shown for reference (note that some roads within 10 km of the center are outside the regional boundary)."}
# nrow(r_central)
# nrow(rj)
square = stplanr::geo_bb(shp = city_centre_buffer)
m1 = tm_shape(r_main_region) +
  tm_lines("maxspeed", palette = "-Set1", lwd = "n_lanes_numeric", scale = 5,
           lwd.legend = c(1, 2, 3, 4, 5, 6), title.lwd = "Number of lanes") +
  tm_shape(region) + tm_borders() +
  tm_shape(square) + tm_borders(lty = 3) +
  tm_shape(h_leeds) + tm_dots(size = 0.5) +
  tm_scale_bar() 
# m1
m2 = tm_shape(r_main) +
  tm_lines("maxspeed", palette = "-Set1", lwd = "n_lanes_numeric", scale = 5,
           lwd.legend = c(1, 2, 3, 4, 5, 6), title.lwd = "Number of lanes") +
  tm_shape(city_centre_buffer) + tm_borders(lty = 2) +
  tm_shape(city_centre_buffer_small) + tm_borders(lty = 2) +
  tm_shape(city_centre_buffer_large) + tm_borders(lty = 2) +
  tm_shape(city_key_buffer) + tm_borders() +
  tm_shape(h_leeds) + tm_dots(size = 0.5) +
  tm_scale_bar() +
  tm_layout(legend.show = FALSE)
tmap_arrange(m1, m2, nrow = 1)
```

>>>>

---
title: "Data analysis to prioritise new cycleways in response to COVID-19"
author: Dr Robin Lovelace & Dr Joey Talbot, Institute for Transport Studies, University of Leeds
output: bookdown::pdf_document2
bibliography: report-refs.bib
---

```{r, eval=FALSE, echo=FALSE}
citr::tidy_bib_file("~/uaf/allrefs.bib", rmd_file = "report.Rmd", file = "report-refs.bib")
citr::tidy_bib_file("article.Rmd", "~/uaf/allrefs.bib", file = "ref.bib")
browseURL("report.pdf")
```

```{r setup2, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, fig.align = "center")
```

# Introduction

New statutory guidance and funding for active transport infrastructure was announced by the Secretary of State for Transport on May 9^th^, to support the UK's response to the COVID-19 crisis by enabling healthy and safe travel at a time of reduced capacity on public transport networks.^[
https://www.gov.uk/government/publications/reallocating-road-space-in-response-to-covid-19-statutory-guidance-for-local-authorities/traffic-management-act-2004-network-management-in-response-to-covid-19
]

The statutory guidance on network management in response to COVID-19 outlines a range of measures that can support a safe and healthy restart, including the creation of new cycleways, reducing speed limits, widening of footways and closing roads to motor traffic via 'modal filters'.
**This report describes the rapid cycleway prioritisation tool project, which provides aims to ensure evidence-based and effective investment in first of these measures: the creation of new cycleways in existing carriageway space reclaimed from motor traffic.**
Light segregation infrastructure such as flexible plastic wands (illustrated in Figure 1) and bollards can deliver a step change in provision for cycling, when implemented in combination with other measures such reduced speed limits on key corridors (as has happened on the A33 in Southampton and the A56 in Manchester) and when placed where they are most needed.
**The purpose of the tool is to help identify such key corridors, ensuring that funds from the Emergency Active Travel Fund change the status quo of city transport systems in favour of active travel.**

<!-- 17:29:01	 From Gill M (Manchester cyclist) : Trafford is planning to reduce speeds on Bridgewater Way element of A56 when the pop up lane goes in there “OneTrafford tweeted that work is continuing to add pop up cycle lanes along Cornbrook Road and Warwick Road in Stretford too, with temporary 30mph speed limit imposed along Bridgewater Way.” Source (https://www.messengernewspapers.co.uk/news/18489349.cyclists-praise-pop-up-cycle-lanes-along-a56-trafford/) -->

<!-- https://twitter.com/doddyuk/status/1267146449227255808 -->

```{r park-lane, fig.cap="New space for cycling created by road space reallocation in Park Lane, London.", out.width="60%"}
# source: Hillingdon and Uxbridge Times
# https://www.hillingdontimes.co.uk/news/national/18450461.one-londons-prestigious-roads-gets-cycle-lane-due-pandemic/
knitr::include_graphics("figures/cycleway-image.jpg")
```

The £250m released by the Emergency Active Travel Fund will be spent in three broad ways:

- The first tranche: £45m released on 1^st^ June, is for swift interventions that must be implemented by the end of June
- The second tranche: £180m will be released later in the summer for more permanent measures
- £25m will be invested in cycle repair schemes

The emphasis of the evidence presented in this report is providing a strong evidence base to prioritisation of schemes eligible for the second tranche.
However, the evidence could also be used to support identification of schemes for the first tranche, such as reallocation of road space on arterial routes, where decisions have not yet been made.

Given the importance of acting quickly, and the unique opportunity that the crisis poses to improve physical activity levels in the long term, it is vital that local authorities have the evidence they need.
Without a nationally consistent and systematic evidence base, there is a risk that some already stretched local authorities will rely on ad-hoc methods to allocate funding.
The rapid cycleway prioritisation tool can also support authorities to make bold plans by highlighting the potential of key strategic corridors.

# Methods

The approach was designed to identify roads and road segments where investment in new cycleways should be prioritised using a transparent methodology and nationally consistent datasets, alongside local authority expertise and local knowledge.
Strong and actionable datasets already exist to support strategic cycle network planning, notably the Propensity to Cycle Tool (PCT), a national planning support system [@lovelace_propensity_2017].
The PCT is already used by local authorities and contains key data showing not only current levels of cycling but also scenarios of growth, notably the Government Target scenario of doubling the number of cycling stages made by 2025 [@departmentfortransport_cycling_2017], at area, route and route network levels, for commuting and school travel (as of May 2020) [@goodman_scenarios_2019].
A limitation of the PCT, however, is that it takes little account of existing infrastructure: variables such as road width and even the presence/absence of cycle infrastructure currently are not included in the results.

To address this limitation, the work set out below will use national road network data which includes geographic attributes and cycle infrastructure from OpenStreetMap, plus carriageway width data derived from analysis of Ordnance Survey data.
This road data was developed for the Cycling Infrastructure Prioritisation Tool (CyIPT) project, funded as a prototype project by a DfT Transport-technology research innovation grants (T-TRIG).
The results of CyIPT, while not publicly accessible, are well known among local authority transport planners and are reported to be of great use to practitioners in scheme prioritisation, according to a DfT funded Discovery project on tools to support investment in active transport.^[
Planning Tools for cycle networks and infrastructure prioritisation, unpublished.
]

The rapid cycleway prioritisation tool is based on levels of cycling potential *and* road attribute data, with an emphasis on road space reallocation.
The method was parameterised to highlight roads that have sufficient space to allow for new cycleways without affecting the navigable network for motorised modes [@lovelace_methods_2020].
A key benefit of this method is that it allows continuous road sections to be identified and ranked, supporting prioritisation of investment, based not only on cycling potential, but also based on continuous road length and estimated cycle usage measured in km per day.
The tool highlights continuous road sections with high cycling potential in the context of existing infrastructure on the basis that new infrastructure should maximise cycle network cohesion and continuity [@parkin_designing_2018].

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE, cache = FALSE, out.width = "100%", eval = FALSE)
```
